Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-04-25 manuscript.tex
latest

For the majority of this manuscript we assume that mortality only occurs in adults (setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
\begin{equation}
\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
\end{equation}
This is reasonable approximation in the absence of disturbances; when we come to consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), we will incorporate disturbance-induced mortality in competing juveniles (Fig.~\ref{fig:lottery}).

@@

The ideal environment is characterized by the near-absence of stress and disturbance: $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$.

@@

Disturbed environments are characterized by short bursts of high elevated extrinsic mortality caused by physical destruction

high extrinsic mortality , or, more accurately, .

In the disturbance extreme of Grime's triangle we can

, we instead represent disturbance by high constant mortality rates $d_i$, and extend this mortality to juveniles in the process of territorial contest (since disturbances do not only affect adults as in Eq. \eqref{eq:delttot}).

We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration. Disturbed environments then correspond to $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration).

@@

Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein). Severe stress inhibits growth and reproduction, so that $b\ll 1$ \citep{grime_1974,grime_1977}. Mutations which appreciably improve $b$ will be either non-existent or extremely unlikely, so $b$ is constrained to remain low. In Grime's view, under these conditions the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate. This implies $b/d\approx 1$ in our model, and so the propagule density $L$ is suppressed to such low levels that there are essentially no territorial contests occurring.

The alternative view is that, while stressful environments imply lower $b$ and support a lower number of individuals per unit area compared what is attainable in ideal environments, stressed populations are actually at high densities relative to the environmental carrying capacity, and are highly competitive \citep{taylor_1990}. In the particular case that stress is caused by scarcity of consumable resources, we might expect intense resource competition (for empirical support, see \citealt{davis_1998}). Thus, $b$ may actually appreciably exceed $d$ under stressful conditions, even though the absolute value of $b$ is small.

@@

How does Fig.~4 compare to empirical analyses of Grime's C/S/R strategies? In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}. The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial.

In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$. Higher $b$ means higher fecundity and/or dispersal ability (see ``Model''). This is consistent with a ruderal strategy. Plant ruderals devote a large proportion of their productivity to seed production \citep{grime_1977}, whereas the analogous ``opportunist'' fishes have large intrinsic growth rates \citep{winemiller_1992}. In corals, ruderals are distinguished by brood spawning (rather than broadcast spawning; \citealt{darling_2012}). This corresponds to higher parental investment and lower overall propagule production --- counter-intuitively, a stereotypical ``$K$-selected'', high-density trait \citep{pianka_1970}. However, since broadcast spawners are vulnerable to an Allee effect at the egg fertilization stage \citep{knowlton_2001}, brood spawning could actually be a way to ensure high $b$ at low densities \citep{darling_2012}. Lower $d$ could be achieved by improved individual resistance to physical destruction, but it is hard to reduce mortality in the face of severe disturbances. Alternatively, shortening the time to reproductive maturity (the iteration time in our model) is an effective way of reducing the chance of death per iteration, $d$, for a given frequency of disturbance. An exceptionally short life cycle is probably the most defining characteristic of ruderals \citep{grime_1977,winemiller_1992,darling_2012}. Note that if evolution manages to appreciably reduce $d$ for a given disturbance intensity, then the population no longer lies at the extreme disturbance vertex of Grime's triangle. Thus, ruderals are characterized by both high $b$ and high $d$, but there is a constant pressure for a shorter life cycle, extending the limits of disturbance that can be tolerated.

In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the high-density interpretation of stressful environments. Low $d$ is essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals). For fishes, the ``equilibrium'' strategy is the analogue of Grime's stress tolerator. This strategy is associated with consumable resource limitation, and is also characterized by long life span, as well as high parental investment in tiny broods. This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the high-density interpretation).

In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$. In plants and corals, a key mechanism for winning territorial contests (higher $c$) is rapidly outgrowing and ``shading out'' competitors; empirically, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}. The situation for $b$ is more ambiguous. Competitor strategies in plants and corals span a range of $b$. For fishes, the analogous ``periodic'' strategy is characterized by enormous spawn sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}. The role of $b$ in ideal environments will be discussed further below.

>>>>

For the majority of this manuscript we assume that mortality only occurs in adults (setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
\begin{equation}
\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
\end{equation}
This seems reasonable in the absence of disturbances; when we come to consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), we will incorporate disturbance-induced mortality in competing juveniles (Fig.~\ref{fig:lottery}).

@@

The ideal environment is characterized by the near-absence of stress and disturbance: $d\ll 1$, whereas $b$ is potentially much larger than $1$. Disturbed environments are characterized by short bursts of high extrinsic mortality $d$ caused by physical destruction. We assume that in the disturbance extreme of Grime's triangle, many such bursts (which could each be be spatially localized) will occur each iteration (i.e. over the time from birth to reproductive maturity), and so we can approximate extreme disturbance by high constant mortality rates $d_i$, which are now extended to affect juveniles as well. We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration. Disturbed environments then correspond to $d_i$ being close to $1$ for all genotypes (most adults and juveniles are killed over an iteration).

@@

@@

Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein). Severe stress inhibits growth and reproduction, so that $b\ll 1$ \citep{grime_1974,grime_1977}. In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$. In our model, this implies that territorial contests are rare.

The alternative view is that stressed environments are highly competitive \citep{taylor_1990}. In particular, if consumable resources are scarce, we expect intense resource competition (for empirical support, see \citealt{davis_1998}). Stress does restrict growth and reproduction, but it also means that fewer individuals can be supported per unit area. Thus, stressed populations are actually at high density relative the environmental carrying capacity (but low density relative to ideal environments).

@@

%How does Fig.~4 compare to empirical analyses of Grime's C/S/R strategies? In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}. The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial. 
%
%In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$. Higher $b$ means higher fecundity and/or dispersal ability (see ``Model''). This is consistent with a ruderal strategy. Plant ruderals devote a large proportion of their productivity to seed production \citep{grime_1977}, whereas the analogous ``opportunist'' fishes have large intrinsic growth rates \citep{winemiller_1992}. In corals, ruderals are distinguished by brood spawning (rather than broadcast spawning; \citealt{darling_2012}). This corresponds to higher parental investment and lower overall propagule production --- counter-intuitively, a stereotypical ``$K$-selected'', high-density trait \citep{pianka_1970}. However, since broadcast spawners are vulnerable to an Allee effect at the egg fertilization stage \citep{knowlton_2001}, brood spawning could actually be a way to ensure high $b$ at low densities \citep{darling_2012}. Lower $d$ could be achieved by improved individual resistance to physical destruction, but it is hard to reduce mortality in the face of severe disturbances. Alternatively, shortening the time to reproductive maturity (the iteration time in our model) is an effective way of reducing the chance of death per iteration, $d$, for a given frequency of disturbance. An exceptionally short life cycle is probably the most defining characteristic of ruderals \citep{grime_1977,winemiller_1992,darling_2012}. Note that if evolution manages to appreciably reduce $d$ for a given disturbance intensity, then the population no longer lies at the extreme disturbance vertex of Grime's triangle. Thus, ruderals are characterized by both high $b$ and high $d$, but there is a constant pressure for a shorter life cycle, extending the limits of disturbance that can be tolerated. 
%
%In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the high-density interpretation of stressful environments. Low $d$ is essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals). For fishes, the ``equilibrium'' strategy is the analogue of Grime's stress tolerator. This strategy is associated with consumable resource limitation, and is also characterized by long life span, as well as high parental investment in tiny broods. This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the high-density interpretation).
%
%In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$. In plants and corals, a key mechanism for winning territorial contests (higher $c$) is rapidly outgrowing and ``shading out'' competitors; empirically, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}. The situation for $b$ is more ambiguous. Competitor strategies in plants and corals span a range of $b$. For fishes, the analogous ``periodic'' strategy is characterized by enormous spawn sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}. The role of $b$ in ideal environments will be discussed further below.
